---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# RPresto

<!-- badges: start -->
<!-- badges: end -->

RPresto is a [DBI](https://dbi.r-dbi.org/)-based adapter for
the open source distributed SQL query engine [Presto](https://prestodb.io/)
for running interactive analytic queries.

## Installation

RPresto is both on [CRAN](https://cran.r-project.org/package=RPresto)
and [github](https://github.com/prestodb/RPresto).

For the CRAN version, you can use

```{r install_cran, eval = FALSE}
install.packages("RPresto")
```

You can install the development version of RPresto from
[GitHub](https://github.com/) with:

```{r install_github, eval = FALSE}
# install.packages("devtools")
devtools::install_github("prestodb/RPresto")
```

## Usage

**The following examples assume that you have a in-memory Presto server set up
locally.** It's the simplest server which stores all data and metadata in RAM on
workers and both are discarded when Presto restarts. If you don't have one set
up, please refer to the [memory connector documentation][1].

```{r setup, eval = TRUE}
# Load libaries and connect to Presto
library(RPresto)
library(DBI)

con <- DBI::dbConnect(
  drv = RPresto::Presto(),
  host = "http://localhost",
  port = 8080,
  user = Sys.getenv("USER"),
  catalog = "memory",
  schema = "default"
)
```

There are two levels of APIs: `DBI` and `dplyr`.

### `DBI` APIs

The easiest and most flexible way of executing a `SELECT` query is using a
[`dbGetQuery()`][2] call. It returns the query result in a [`tibble`][3].

```{r dbGetQuery, eval=TRUE}
DBI::dbGetQuery(con, "SELECT CAST(3.14 AS DOUBLE) AS pi")
```

[`dbWriteTable()`][4] can be used to write a small data frame into a Presto
table.

```{r dbWriteTable_pre_hook, eval = TRUE, echo = FALSE}
if (DBI::dbExistsTable(con, "iris")) {
  DBI::dbRemoveTable(con, "iris")
}
```

```{r dbWriteTable, eval = TRUE}
# Writing iris data frame into Presto
DBI::dbWriteTable(con, "iris", iris)
```

[`dbExistsTable()`][5] checks if a table exists.

```{r dbExistsTable, eval = TRUE}
DBI::dbExistsTable(con, "iris")
```

[`dbReadTable()`][6] reads the entire table into R. It's essentially a `SELECT *`
query on the table.

```{r dbReadTable, eval = TRUE}
DBI::dbReadTable(con, "iris")
```

[`dbRemoveTable()`][7] drops the table from Presto.

```{r dbRemoveTable, eval = TRUE}
DBI::dbRemoveTable(con, "iris")
```

You can execute a statement and returns the number of rows affected using
[`dbExecute()`][8].

```{r dbExecute_1_pre_hook, eval = TRUE, echo = FALSE}
if (DBI::dbExistsTable(con, "testing_table")) {
  DBI::dbRemoveTable(con, "testing_table")
}
```

```{r dbExecute_1, eval = TRUE}
# Create an empty table using CREATE TABLE
DBI::dbExecute(
  con, "CREATE TABLE testing_table (field1 BIGINT, field2 VARCHAR)"
)
```

`dbExecute()` returns the number of rows affected by the statement. Since a
`CREATE TABLE` statement creates an empty table, it returns 0.

```{r dbExecute_2, eval = TRUE}
DBI::dbExecute(
  con,
  "INSERT INTO testing_table VALUES (1, 'abc'), (2, 'xyz')"
)
```

Since 2 rows are inserted into the table, it returns 2.

```{r check_dbExecute, eval = TRUE}
# Check the previous INSERT statment works
DBI::dbReadTable(con, "testing_table")
```

### `dplyr` APIs

We also include `dplyr` database backend integration (which is mainly
implemented using the [`dbplyr` package][9]).

```{r dplyr_setup, eval = TRUE, message = FALSE}
# Load packages
library(dplyr)
library(dbplyr)
```

We can use [`dplyr::copy_to()`][11] to write a local data frame to a
`PrestoConnection` and immediately create a remote table on it.

```{r dplyr_copyto, eval = TRUE}
# Add mtcars to Presto
if (DBI::dbExistsTable(con, "mtcars")) {
  DBI::dbRemoveTable(con, "mtcars")
}
tbl.mtcars <- dplyr::copy_to(dest = con, df = mtcars, name = "mtcars")
# colnames() gives the column names
tbl.mtcars %>% colnames()
```

[`dplyr::tbl()`][10] also work directly on the `PrestoConnection`.

```{r dplyr_tbl, eval = TRUE}
# Treat "iris" in Presto as a remote data source that dplyr can now manipulate
if (!DBI::dbExistsTable(con, "iris")) {
  DBI::dbWriteTable(con, "iris", iris)
}
tbl.iris <- dplyr::tbl(con, "iris")

tbl.iris %>% colnames()

# dplyr verbs can be applied onto the remote data source
tbl.iris %>%
  group_by(species) %>%
  summarize(
    mean_sepal_length = mean(sepal.length, na.rm = TRUE)
  ) %>%
  arrange(species) %>%
  collect()
```

## Connecting to Trino

To connect to Trino you must set the `use.trino.headers` parameter so `RPresto`
knows to send the correct headers to the server. Otherwise all the same
functionality is supported.

```{r, eval=FALSE}
con.trino <- DBI::dbConnect(
  RPresto::Presto(),
  use.trino.headers=TRUE,
  host="http://localhost",
  port=7777,
  user=Sys.getenv("USER"),
  schema="<schema>",
  catalog="<catalog>",
  source="<source>"
)
```

## Passing extra credentials to the connector

To pass extraCredentials that gets added to the `X-Presto-Extra-Credential`
header use the `extra.credentials` parameter so `RPresto` will add that to the
header while creating the `PrestoConnection`.

Set `use.trino.headers` if you want to pass extraCredentials through the
`X-Trino-Extra-Credential` header.

```{r, eval=FALSE}
con <- DBI::dbConnect(
  RPresto::Presto(),
  host="http://localhost",
  port=7777,
  user=Sys.getenv("USER"),
  schema="<schema>",
  catalog="<catalog>",
  source="<source>",
  extra.credentials="test.token.foo=bar",
)
```

## How RPresto works

Presto exposes its interface via a REST based API[^1]. We utilize the
[httr](https://github.com/r-lib/httr) package to make the API calls and
use [jsonlite](https://github.com/jeroen/jsonlite) to reshape the
data into a `tibble`. Note that as of now, only read operations are
supported.

RPresto has been tested on Presto 0.100.

## License
RPresto is BSD-licensed.

[^1]: See <https://github.com/prestodb/presto/wiki/HTTP-Protocol> for a
description of the API.

[1]: https://prestodb.io/docs/current/connector/memory.html
[2]: https://dbi.r-dbi.org/reference/dbgetquery
[3]: https://tibble.tidyverse.org/
[4]: https://dbi.r-dbi.org/reference/dbwritetable
[5]: https://dbi.r-dbi.org/reference/dbexiststable
[6]: https://dbi.r-dbi.org/reference/dbreadtable
[7]: https://dbi.r-dbi.org/reference/dbremovetable
[8]: https://dbi.r-dbi.org/reference/dbexecute
[9]: https://dbplyr.tidyverse.org/
[10]: https://dplyr.tidyverse.org/reference/tbl.html
[11]: https://dplyr.tidyverse.org/reference/copy_to.html
