---
title: "Translating complex Presto data types"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Translating complex Presto data types}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = RPresto::presto_has_default()
)
```

In this vignette, we demonstrate how complex structural types in Presto can be
translated into R types (e.g., vectors, list, and tibbles).

## Package setup

```{r setup}
library(RPresto)
```

You can check your `RPresto` version by running the `packageVersion()` function.
**You need version 1.3.9 or later to have a more comprehensive and robust
complex types support.**

```{r version}
packageVersion("RPresto")
```

## Overview

Complex types refer to [structural types][1] including `ARRAY`, `MAP` and `ROW`.
Those data types are basically containers to hold other data types (hence
*complex*).

We summarize the similarities and differences between the complex types below.

| Type | Is atomic? | Has names/keys? |
|------|:----------:|:---------------:|
| ARRAY | Yes | No |
| MAP | Yes | Yes |
| ROW | No | Yes |

Atomic here means all elements in the container share the same type (usually
primitive types, but can be complex types too). For example, an `ARRAY` of
integer can only hold integer type elements whereas a `ROW` can have elements of
different types (e.g., one element is integer and the other boolean).

## Walkthrough preparation

### Local Presto server in memory

We assume that the user already have a Presto server with a memory connector set
up. If you don't have such a server set up, refer to the
[Presto documentation][2] for instructions if you want to follow along.

### Presto connection in R

We first create a `PrestoConnection` which will serve as the bridge between the
Presto server and R.

```{r PrestoConnection}
con <- DBI::dbConnect(
  drv = RPresto::Presto(),
  host = "http://localhost",
  port = 8080,
  user = Sys.getenv("USER"),
  catalog = "memory",
  schema = "default"
)
```

We first issue a simple query to see if the Presto connection is working
properly.

```{r test}
DBI::dbGetQuery(con, "SELECT 1+1 AS res")
```

## Atomic structural types that contain primitive types

| Presto type | R type |
|-------------|:------:|
| ARRAY | unnamed typed vector |
| MAP | named typed vector |

### `ARRAY`s

Given the atomic and unnamed nature of `ARRAY`s, we map them to **unnamed typed
vectors** in R.

#### Create a table with `ARRAY`s of primitive types

We first create a table with `ARRAY`s of all supported primitive Presto data
types using the `create_primitive_arrays_table()` function included
in the `RPresto` package.

```{r primitive-arrays-table}
RPresto:::create_primitive_arrays_table(
  con, table_name = "presto_primitive_arrays", verbose = FALSE
)
```

We can check if the table now exists in Presto.

```{r check-primitive_arrays-table}
DBI::dbExistsTable(con, "presto_primitive_arrays")
```

We can list the fields in the table. They are named after the Presto types they
represent.

```{r primitive-arrays-table-fields}
DBI::dbListFields(con, "presto_primitive_arrays")
```

#### Translate `ARRAY`s to R vectors

```{r array-of-primitive-types}
(
  df.array_of_primitive_types <- dbGetQuery(
    con,
    "SELECT * FROM presto_primitive_arrays",
    bigint = "integer64"
  )
)
```

We can verify the R types of each column.

```{r array-of-primitive-type-check}
tibble::enframe(purrr::map_chr(df.array_of_primitive_types, ~class(.[[1]])[1]))
```

All vectors are unnamed.

```{r array-of-primitive-name-check}
purrr::every(df.array_of_primitive_types, ~is.null(names(.[[1]])))
```

We can also call functions such as `length()` on each of the columns to get the
`ARRAY` cardinality. It shows that all `ARRAY`s have 3 elements in them.

```{r array-cardinality}
tibble::enframe(purrr::map_int(df.array_of_primitive_types, ~length(.[[1]])))
```

### `MAP`s

A `MAP` in Presto can be thought as a combination of two same-length `ARRAY`s.
The first `ARRAY` contains the keys of the `MAP` and the second contains the
values of the `MAP`. In fact, that's exactly how `MAP` literals are created in
Presto (e.g., `MAP(ARRAY[1, 2], ARRAY['a', 'b'])` creates a 2-element `MAP`).

Following the logic, we translate `MAP`s to **named typed vectors** in R.

#### Create a table with `MAP`S of primitive types

We first create a table with `MAPS`s of all supported primitive Presto data
types using the `create_primitive_maps_table()` function included
in the `RPresto` package.

```{r primitive-maps-table}
RPresto:::create_primitive_maps_table(
  con, table_name = "presto_primitive_maps", verbose = FALSE
)
```

We can check if the table now exists in Presto.

```{r check-primitive_maps-table}
DBI::dbExistsTable(con, "presto_primitive_maps")
```

We can list the fields in the table. They are named after the Presto types they
represent.

```{r primitive-maps-table-fields}
DBI::dbListFields(con, "presto_primitive_maps")
```

#### Translate `MAP`s to R vectors

```{r map-of-primitive-types}
(
  df.map_of_primitive_types <- dbGetQuery(
    con,
    "SELECT * FROM presto_primitive_maps",
    bigint = "integer64"
  )
)
```

We can verify the R types of each column.

```{r map-of-primitive-type-check}
tibble::enframe(purrr::map_chr(df.map_of_primitive_types, ~class(.[[1]])[1]))
```

All vectors are named.

```{r map-of-primitive-name-check}
purrr::none(df.map_of_primitive_types, ~is.null(names(.[[1]])))
```

### Repeated `ARRAY`s and `MAP`s

It's possible to have repeated `ARRAY`s and `MAP`s in Presto in the form of
`ARRAY`s of `ARRAY`s and `ARRAY`s of `MAP`s.

| Repeated Presto type | R type |
|----------------------|--------|
| ARRAY of ARRAY | **not supported** |
| ARRAY of MAP | unnamed list of named typed vectors |

We are not supporting nested `ARRAY`s at the moment although it's technically
possible in Presto.

For `ARRAY`s of `MAP`s, we translate the `ARRAY` container into an unnamed list
and each of the `MAP` element into a named typed vector.

We first create an **array-of-maps** table by using the
`create_array_of_maps_table()` function.

```{r array-of-maps-table}
RPresto:::create_array_of_maps_table(
  con, table_name = "presto_array_of_maps", verbose = FALSE
)
```

We can check if the table now exists in Presto.

```{r check-array-of-maps-table}
DBI::dbExistsTable(con, "presto_array_of_maps")
```

We can list the fields in the table.

```{r array-of-maps-table-fields}
DBI::dbListFields(con, "presto_array_of_maps")
```

Let's import all the data into R.

```{r array-of-maps}
(
  df.array_of_maps <- dbGetQuery(
    con,
    "SELECT * FROM presto_primitive_maps",
    bigint = "integer64"
  )
)
```

We need to pry open the wrapping unnamed list to reveal the types of the vectors
underneath.

```{r array-of-map-check}
tibble::enframe(purrr::map_chr(df.array_of_maps, ~class(.[[1]][[1]])[1]))
```

## `ROW` type

The easiest way to think about the `ROW` type in Presto is to think of it
literally as a *row* of a table. Just as a table can have multiple columns of
different data types, a `ROW` can have multiple elements of different types. And
just like a table having a name for each column, every element of a `ROW` has a
name associated with the value.

Depending on whether the `ROW` type is repeated (i.e., wrapped in an `ARRAY`),
the translation into R is different.

* We translate single `ROW` value to a named list in R.

* Rather than interpret repeated `ROW`s (i.e., `ARRAY` of `ROW`s) as a list of
named lists, we translate the collection of `ROW`s into a tibble.

| Presto type | R type |
|-------------|--------|
| Single ROW | named list |
| Repeated ROWs | tibble |

### Single `ROW` translation

To demonstrate how `ROW` types are translated into R types, we first create a
table using an auxiliary `create_primitive_rows_table()` function
included in the package. The resulting table has only 1 column named
`row_primitive_types` which is a `ROW` that includes 18 sub-columns representing
all supported primitive types.

```{r primitive-rows-table}
RPresto:::create_primitive_rows_table(
  con, table_name = "presto_primitive_rows", verbose = FALSE
)
```

We can check if the table now exists in Presto.

```{r check-primitive_rows-table}
DBI::dbExistsTable(con, "presto_primitive_rows")
```

We can list the fields in the table.

```{r primitive-rows-table-fields}
DBI::dbListFields(con, "presto_primitive_rows")
```

We can then retrieve all the data from the table.

```{r row-of-primitive}
(
  df.row_of_primitive <- dbGetQuery(
    con,
    "SELECT row_primitive_types FROM presto_primitive_rows",
    bigint = "integer64"
  )
)
```

We can check the R types of each element in the named list.

```{r row-of-primitive-check}
tibble::enframe(
  purrr::map_chr(df.row_of_primitive$row_primitive_types[[1]], ~class(.)[1])
)
```

### Repeated `ROW`s translation

To construct a repeated `ROW` column, we use the auxiliary
`create_array_of_rows_table()` function.

```{r arrays-of-rows-table}
RPresto:::create_array_of_rows_table(
  con, table_name = "presto_array_of_rows", verbose = FALSE
)
```

We can check if the table now exists in Presto and the field name.

```{r check-array-of-rows-table}
DBI::dbExistsTable(con, "presto_array_of_rows")
```

```{r array-of-rows-table-fields}
DBI::dbListFields(con, "presto_array_of_rows")
```

We can import the whole data into R.

```{r array-of-rows}
(
  df.array_of_rows <- dbGetQuery(
    con,
    "SELECT array_of_rows FROM presto_array_of_rows",
    bigint = "integer64"
  )
)
```

We can verify the tibble's column types.

```{r array-of-row-check}
tibble::enframe(
  purrr::map_chr(df.array_of_rows$array_of_rows[[1]], ~class(.)[1])
)
```

## Walkthrough wrap-up

```{r close-db}
DBI::dbDisconnect(con)
```

[1]: https://prestodb.io/docs/current/language/types.html#structural
[2]: https://prestodb.io/docs/current/connector/memory.html
