---
title: "Primitive Presto data types to R types"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Primitive Presto data types to R types}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = RPresto::presto_has_default()
)
options(pillar.max_dec_width=20)
```

In this vignette, we introduce how primitive Presto data types are translated
into R types in the `RPresto` package.

## Package setup

```{r setup}
library(RPresto)
```

You can check your `RPresto` version by running the `packageVersion()` function.
**You need version 1.3.9 or later to have a more comprehensive and robust
primitive types support.**

```{r version}
packageVersion("RPresto")
```

## Overview

Here *primitive types* refer to basic data types that are atomic and non-nested
(as opposed to structural types such as `ARRAY`, `MAP`, and `ROW`). Refer to the
[Presto documentation][1] for
a complete list of Presto data types.

We don't currently support all of Presto's primitive types. Here we summarize
what's supported and what those supported types map to in R.

| Category | Presto data type | R type |
|----------|------------------|--------|
| Boolean | BOOLEAN | logical |
| Integer | TINYINT | integer |
| Integer | SMALLINT | integer |
| Integer | INTEGER | integer |
| Integer | BIGINT | \{**integer**\|integer64\|numeric\|character\} |
| Floating-point | REAL | numeric |
| Floating-point | DOUBLE | numeric |
| Floating-point | DECIMAL | **not supported** |
| String | VARCHAR | character |
| String | CHAR | character |
| String | VARBINARY | raw |
| String | JSON | **not supported** |
| Date and Time | DATE | Date (S3 class) |
| Date and Time | TIME | difftime (S3 class) |
| Date and Time | TIME WITH TIME ZONE | difftime (S3 class) |
| Date and Time | TIMESTAMP | POSIXct (S3 class) |
| Date and Time | TIMESTAMP WITH TIME ZONE | POSIXct (S3 class) |
| Date and Time | INTERVAL YEAR TO MONTH | Duration (S4 class) |
| Date and Time | INTERVAL DAY TO SECOND | Duration (S4 class) |

## Walkthrough preparation

### Local Presto server in memory

We assume that the user already have a Presto server with a memory connector set
up. If you don't have such a server set up, refer to the
[Presto documentation][2] for instructions if you want to follow along.

### Presto connection in R

We first create a `PrestoConnection` which will serve as the bridge between the
Presto server and R.

```{r PrestoConnection}
con <- DBI::dbConnect(
  drv = RPresto::Presto(),
  host = "http://localhost",
  port = 8080,
  user = Sys.getenv("USER"),
  catalog = "memory",
  schema = "default"
)
```

We first issue a simple query to see if the Presto connection is working
properly.

```{r test}
DBI::dbGetQuery(con, "SELECT 1+1 AS res")
```

### Primitive data types table in Presto

To illustrate how those primitive types are mapped to R types in bulk, we first
create a made-up table with 3 rows and 17 fields (i.e., one column for each
supported Presto type).

We create the table using an auxiliary
`create_primitive_types_table()` function included in the package.

```{r primitive-types-table}
RPresto:::create_primitive_types_table(
  con, table_name = "presto_primitive_types", verbose = FALSE
)
```

We can check if the table now exists in Presto.

```{r check-primitive_types-table}
DBI::dbExistsTable(con, "presto_primitive_types")
```

We can list the fields in the table. They are named after the Presto types they
represent.

```{r primitive-types-table-fields}
DBI::dbListFields(con, "presto_primitive_types")
```

## Data types mapping to R

### Boolean

Translating boolean values from Presto to R is fairly straightforward. `true`
and `false` values are mapped to `TRUE` and `FALSE` in R and `null` is mapped to
`NA` which is by default a `logical` (i.e., boolean) type in R.

```{r boolean-type}
(
  df.boolean <- dbGetQuery(con, "SELECT boolean FROM presto_primitive_types")
)
```

We can verify that all three values in R are `logical`.

```{r boolean-check}
purrr::map_chr(df.boolean$boolean, class)
```

### Integers

Presto has 4 integer data types.

* [`TINYINT`][3] is 8-bit and ranges from `-2^7` to `2^7-1`.

* [`SMALLINT`][4] is 16-bit and ranges from `-2^15` to `2^15-1`.

* [`INTEGER`][5] is 32-bit and ranges from `-2^31` to `2^31-1`.

* [`BIGINT`][6] is 64-bit and ranges from `-2^63` to `2^63-1`.

| Presto data type | Bits | Minimum value | Maximum value |
|------------------|:----:|:-------------:|:-------------:|
| TINYINT | 8 | -128 | 127 |
| SMALLINT | 16 | -32,768 | 32,767 |
| INTEGER | 32 | -2,147,483,648 | 2,147,483,647 |
| BIGINT | 64 | -9,223,372,036,854,775,808 | 9,223,372,036,854,775,807 |

In comparison, base R only ships with one 32-bit integer type, but the range is
slightly different from Presto. R's integer type ranges from
`-2^31+1 = -2,147,483,647` to `2^31-1 = 2,147,483,647`. The one number (out of
~4.3 billion 32-bit integer numbers) that's a valid `INTEGER` in Presto but not
in R is the range lower bound `-2,147,483,648`.

#### Non-BIGINT integers

Since all `TINYINT`, `SMALLINT`, and `INTEGER` numbers (except `-2,147,483,648`)
can be fit into R's `integer` type, we translate all of them to `integer`s.

```{r non-bigint-type}
(
  df.non_bigint_int <- dbGetQuery(
    con, "SELECT tinyint, smallint, integer FROM presto_primitive_types"
  )
)
```
We can verify that all three columns in R are `integer` types.

```{r non-bigint-check}
purrr::map_chr(df.non_bigint_int, class)
```

We can also show that mapping the one exception `-2,147,483,648` to R results in
an error while the query runs correctly in Presto. **Users whose data contains
the exceptional integer should consider using the `BIGINT` type rather than
`INTEGER` to encode the data in Presto.**

```{r non-bigint-exception, error=TRUE}
dbGetQuery(con, "SELECT CAST('-2147483648' AS INTEGER) AS non_bigint_exception")
```

#### BIGINT integers

Since the Presto `BIGINT` type is 64-bit, its range is clearly much wider than
base R's `integer` type.

However, by default we still map `BIGINT` integers from Presto to `integer` in
R, for two reasons.

1. R doesn't have native 64-bit integer support yet. Mapping `BIGINT` to a
64-bit integer in R will require additional package and add extra dependencies.

1. The 32-bit integer range covers 2+ billion positive integers and 4.3 billion
integers in total. Most integer values used in data analysis and modeling can be
well covered in that range.

Besides the default mapping to `integer` (albeit 32-bit) type, we offer three
other options when it comes `BIGINT` mapping.

1. Firstly, using the `bigint = "character"` argument instructs `RPresto` to
cast the `BIGINT` value to a `character` type. This is particularly useful when
`BIGINT` is used to store long IDs rather than large numbers (i.e., the numbers
are not used in arithmetic computations).

1. Using `bigint = "integer64"` makes `RPresto` translate `BIGINT` values to a
value of the `integer64` S3 class from the `bit64` package.

1. `bigint = "numeric"` makes `RPresto` store the `BIGINT` value in a `numeric`
( i.e., `double`) type.

Whenever we map an integer to another numeric type, we always need to consider
the precision of the mapping, that is, whether there's any precision loss during
the translation. Among the three translation options above, the first one (i.e.,
casting `BIGINT` to `character`) doesn't involve any precision translation, so
we will focus the precision discussion on the other two translations (
`bit64::integer64` and `numeric`).

On the receiving end, the `bit64::integer64` type has a range from `-2^63+1 = -9,223,372,036,854,775,807` to `2^63-1 = 9,223,372,036,854,775,807`.

```{r integer64-limits}
bit64::lim.integer64()
```

Comparing the range to Presto's `BIGINT` range, it seems that in theory the
`bit64::integer64` range is only infinitesimally smaller than the `BIGINT` range
by 1 number (again, the lower bound number). However, in practice, the range of
`BIGINT` values that can be translated into `bit64::integer64` **without
precision loss** is much smaller.

The limitation comes from how the [Presto REST API][7] communicates data with R.
It uses the JSON format to encode the query result data and sends it to R for
`RPresto` to process.

JSON by default encodes integers as `double` numbers, so its [precision is
limited][8] to `+/-(2^53-1) = +/-9,007,199,254,740,991`. Any integers outside of
this range will lose precision during the translation.

Since the translation limitation is caused by the JSON format encoding integers
using `double`, the same limitation applies when `BIGINT` values are mapped to
`numeric` types in R.

| bigint = | R type | Range without precision loss | Range with possible precision loss |
|----------|--------|------------------------------|------------------------------------|
| character | character | NA | NA |
| integer64 | bit64::integer64 | +/-9,007,199,254,740,991 | +/- 9,223,372,036,854,775,807 |
| numeric | numeric | +/-9,007,199,254,740,991 | [-9,223,372,036,854,775,808, 9,223,372,036,854,775,807] |

The table created by the `create_primitive_types_table()` function
has the values from the no-precision-lost range. Below we show how different
`bigint` arguments change the output R types.

* `bigint = "character"`

```{r bigint-character}
dbGetQuery(
  con, "SELECT bigint FROM presto_primitive_types", bigint = "character"
)
```

* `bigint = "integer64"`

```{r bigint-integer64}
dbGetQuery(
  con, "SELECT bigint FROM presto_primitive_types", bigint = "integer64"
)
```

* `bigint = "numeric"`

```{r bigint-numeric}
dbGetQuery(
  con, "SELECT bigint FROM presto_primitive_types", bigint = "numeric"
)
```

If you attempt to translate integers outside of the no-precision-loss range, a
warning message will show up.

```{r bigint-out-of-range}
dbGetQuery(
  con,
  "
  SELECT SIGN(bigint) * (ABS(bigint) + 1) AS bigint_precision_loss
  FROM presto_primitive_types
  ",
  bigint = "numeric"
)
```

### Floating numbers

R only has one floating point data type, `double` (its class is `numeric`). All
floating numbers are stored in double precision format (i.e., 64-bit). This
matches Presto's [`DOUBLE`][9] type, so translation between Presto and R is
straightforward.

```{r floating-point-type}
(
  df.floating_point <- dbGetQuery(
    con,
    "SELECT real, double FROM presto_primitive_types"
  )
)
```

We can verify that both floating point types are translated to `numeric`.

```{r floating-point-check}
purrr::map_chr(df.floating_point, class)
```

### Strings

#### `VARCHAR` and `CHAR` types

[`VARCHAR`][10] and [`CHAR`][11] data types in Presto are mapped to R's
`character` type.

```{r character-types}
(
  df.characters <- dbGetQuery(
    con,
    "SELECT varchar, char FROM presto_primitive_types"
  )
)
```

We can verify the resulting R types to be `character`.

```{r character-check}
purrr::map_chr(df.characters, class)
```

#### `VARBINARY`

Presto's [`VARBINARY`][12] type stores string data in raw bytes. It can be
nicely mapped to R's `raw` type.

```{r bytes-type}
(
  df.bytes <- dbGetQuery(
    con,
    "SELECT varbinary FROM presto_primitive_types"
  )
)
```

We can verify all elements in the resulting column are of `raw` data type.

```{r bytes-check}
purrr::map_chr(df.bytes$varbinary, class)
```

We can also convert the bytes data back to the string values.

```{r bytes-to-char}
dplyr::mutate(df.bytes, string = purrr::map_chr(varbinary, rawToChar))
```

### Date and Time

In R, there are three types of date/time data that refer to an instant in time.

* A **date**. Tibbles print this as `<date>`.

* A **time** within a day. Tibbles print this as `<time>`.

* A **date-time** is a date plus a time: it uniquely identifies an instant in
time (typically to the nearest second). Tibbles print this as `<dttm>`.
Elsewhere in R these are called `POSIXct`.

We prefer to use the `lubridate` package to handle date and date-time objects.
Time objects are not very commonly used and R lacks a strong native support for
it. In `RPresto`, we uses the `hms` package to handle time objects.

#### `DATE`

The [`DATE`][13] type is by far the most used date-and-time types. Note that a
`DATE` value isn't tied to any particular time zone, so it isn't associated with
a unique point in time (i.e., seconds or microseconds since epoch).

We use base R's `Date` S3 class to translate Presto's `DATE` type.

```{r date-type}
(
  df.date <- dbGetQuery(
    con,
    "SELECT date FROM presto_primitive_types"
  )
)
```

We can verify the R type of the column.

```{r date-check}
purrr::map_chr(df.date, class)
```

#### `TIMESTAMP`

The `POSIXct` type values, on the other hand, are associated with a unique point
in time. That is, they can be translated to a unique numeric value that refers
to the number of time units (usually seconds, milliseconds, or microseconds)
elapsed since epoch (i.e., the beginning of time). This is why the mode of a
`POSIXct` value in R is numeric and you can call the `as.integer()` function on
it.

```{r posixct-value}
foo <- lubridate::ymd_hms("2000-01-01 01:02:03", tz = "America/New_York")
mode(foo)
as.integer(foo)
```

In this case, the integer value refers to the number of seconds elapsed since
1970-01-01 00:00:00 in the UTC time zone.

Presto's [`TIMESTAMP`][14] and [`TIMESTAMP WITH TIME ZONE`][15] types follow the
same logic and thus can be mapped to the `POSIXct` type. To ensure consistency,
`RPresto` always translates the timestamp to match the Presto session's
timezone.

You can check the session time zone by printing the `session.timezone` slot of
the `PrestoConnection` object.

```{r session-tz}
con@session.timezone
```

Here we get the `TIMESTAMP` and `TIMESTAMP WITH TIME ZONE` values from the
table.

```{r posixct-type}
(
  df.posixct <- dbGetQuery(
    con,
    "SELECT timestamp, timestamp_with_tz FROM presto_primitive_types"
  )
)
```

We can check the R types of the two columns.

```{r posixct-check}
purrr::map(df.posixct, class)
```

We can also verify that the time zone attributes match the session time zone.

```{r posixct-tz}
purrr::map_chr(df.posixct$timestamp, lubridate::tz)
purrr::map_chr(df.posixct$timestamp_with_tz, lubridate::tz)
```

#### `INTERVAL`s

Presto has two `INTERVAL` types: [`INTERVAL YEAR TO MONTH`][16] and [`INTERVAL
DAY TO SECOND`][17]. We map both of them to `lubridate::Duration-class`.

```{r duration-type}
(
  df.duration <- dbGetQuery(
    con,
    "
    SELECT
      interval_year_to_month,
      interval_day_to_second
    FROM presto_primitive_types
    "
  )
)
```

## Walkthrough wrap-up

```{r close-db}
DBI::dbDisconnect(con)
```

[1]: https://prestodb.io/docs/current/language/types.html
[2]: https://prestodb.io/docs/current/connector/memory.html
[3]: https://prestodb.io/docs/current/language/types.html#tinyint
[4]: https://prestodb.io/docs/current/language/types.html#smallint
[5]: https://prestodb.io/docs/current/language/types.html#integer
[6]: https://prestodb.io/docs/current/language/types.html#bigint
[7]: https://prestodb.io/docs/current/develop/client-protocol.html
[8]: https://developer.mozilla.org/en-US/docs/Web/JavaScript/Reference/Global_Objects/Number/MAX_SAFE_INTEGER
[9]: https://prestodb.io/docs/current/language/types.html#double
[10]: https://prestodb.io/docs/current/language/types.html#varchar
[11]: https://prestodb.io/docs/current/language/types.html#char
[12]: https://prestodb.io/docs/current/language/types.html#varbinary
[13]: https://prestodb.io/docs/current/language/types.html#date
[14]: https://prestodb.io/docs/current/language/types.html#timestamp
[15]: https://prestodb.io/docs/current/language/types.html#timestamp-with-time-zone
[16]: https://prestodb.io/docs/current/language/types.html#interval-year-to-month
[17]: https://prestodb.io/docs/current/language/types.html#interval-day-to-second
